---
title: "pregnancy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pregnancy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

***This vignette is still a draft***.

```{r setup}
library(pregnancy)
```

This is an R package for calculating dates and tracking medications during pregnancy and fertility treatment. 
It extends a private, personal package that I wrote for myself when I was pregnant.
It contains functions and features that I found useful at the time, 
and others that I added when adapting the package for general use.

The functionality goes beyond what's offered by online pregnancy calculators and apps, 
plus there are no concerns (unlike with these sites and apps) about data privacy, tracking or advertising.

## Date calculations

The `calculate_due_date()` function estimates the pregnancy due date. 
The `start_date` is interpreted differently, depending on the `start_type`. 
By default, the `start_type` is the last menstrual period, and the `start_date` is the date this started.
Other `start_date` options, like various transfer days, are useful for those using IVF.

```{r}
# invisibly returns a Date object with the estimated due date
due_date <- calculate_due_date(as.Date("2024-12-10"))
```

Once a due date is know, `how_far()` tells you how far along the pregnancy is on a given date.
It defaults to providing this information for the current date, 
but an alternative `on_date` can be provided. 
This vignette was built on **`r format(Sys.Date(), '%B %d, %Y')`**, 
so for the purposes of reading this article, that counts as "today".

```{r}
how_far(due_date = due_date)
how_far(on_date = as.Date("2025-04-02"), due_date = as.Date("2025-09-16"))
```

The `date_when()` function gives the date when a certain week of pregnancy will be (or was) reached,
and the duration of that date from today:

```{r}
date_when(28, due_date = due_date)
```

By default, the output messages from `how_far()` and `date_when()` are in the second person, i.e. addressed to "you",
but there is also an option to specify another person.
A value of `1` or `"I"` means you'll be addressed in the first person:

```{r}
how_far(due_date = due_date, person = 1)
```

Any other character string will be interpreted as a third-person name,
which is useful, for example, if you're following the pregnancy progress of a partner:


```{r}
date_when(6, due_date = due_date, person = "Ruth")
```

## Tracking medications

*Write this section!*

## Global options

It would be very tedious to have to enter a due date every time you call `how_far()` or `date_when()` over the course of a pregnancy, especially since that date is constant throughout. 
Similarly for the medications table in `medications_remaining()`.
To avoid this, the pregnancy package makes use of **global options**, 
which can be set with the `set_*` family of functions ((`set_due_date()`, `set_person()`, `set_medications()`), 
retrieved with the `get_*` family of functions.
Any global option can be unset by calling its `set_*` function with the argument `NULL`.

### `pregnancy.due_date`

You can set the `pregnancy.due_date` option using the `set_due_date()` function.
When that option is set, if the `due_date` argument to `how_far()` or `date_when()` is `NULL` (the default),
that option is retrieved. 

`set_due_date()` only sets the due date for the current R session. 
To make this option persist (recommended), it can be added to your `.RProfile`. 
The console output of `set_due_date()` provides guidance on how to do this.

```{r}
set_due_date(due_date)
```

Then `how_far()` can be called without needing to specify any arguments, 
and `date_when()` only needs the target week:

```{r}
how_far()
```

To check what the option is set to, use `get_due_date()`.

### `pregnancy.person`

`set_person()` sets the global option `pregnancy.person`. 
If this option is set it will be retrieved to specify the value of `person` in `how_far()` and `date_when()`,
unless it is overriden by passing another value directly to the `person` argument. 
If the option is not set, the default second-person will be used.

```{r}
set_person(1)
how_far()
set_person(NULL)
how_far()
```

### `pregnancy.medications`

*Write this section!*

### An example `.RProfile`

Taking the advice to specify the options in your `.RProfile`, 
here's an example of what that might look like:

``` r
options(
  pregnancy.due_date = as.Date("2025-09-16"),
  pregnancy.person = "I", # addressed in first person
  pregnancy.medications = dplyr::tribble(
    ~medication, ~format, ~quantity, ~start_date, ~stop_date,
    "progynova", "tablet", 3, as.Date("2025-04-21"), as.Date("2025-04-30"),
    "progynova", "tablet", 6, as.Date("2025-05-01"), as.Date("2025-07-11"),
    "cyclogest", "pessary", 2, as.Date("2025-05-03"), as.Date("2025-07-11"),
    "clexane", "injection", 1, as.Date("2025-05-08"), as.Date("2025-09-05")
  )
)
```

