---
title: "pregnancy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pregnancy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(pregnancy)
```

This is an R package for calculating dates and tracking medications during pregnancy and fertility treatment. It extends a private, personal package that I wrote for myself when I was pregnant. It contains functions and features that I found useful at the time, and others that I added when adapting the package for general use.

The functionality goes beyond what's offered by online pregnancy calculators and apps, plus there are no concerns (unlike with these sites and apps) about data privacy, tracking or advertising.

## Date calculations

The `calculate_due_date()` function estimates the pregnancy due date. The `start_date` is interpreted differently, depending on the `start_type`. By default, the `start_type` is the last menstrual period, and the `start_date` is the date this started. Other `start_date` options, like various transfer days, are useful for those using IVF.

```{r}
# invisibly returns a Date object with the estimated due date
due_date <- calculate_due_date(as.Date("2025-02-24"))
```

Once a due date is know, `how_far()` tells you how far along the pregnancy is on a given date. It defaults to providing this information for the current date, but an alternative `on_date` can be provided. This vignette was built on **`r format(Sys.Date(), '%B %d, %Y')`**, so for the purposes of reading this article, that counts as "today".

```{r}
how_far(due_date = due_date)
how_far(on_date = as.Date("2025-09-17"), due_date = due_date)
```

The `date_when()` function gives the date when a certain week of pregnancy will be (or was) reached, and the duration of that date from today:

```{r}
date_when(33, due_date = due_date)
```

By default, the output messages from `how_far()` and `date_when()` are in the second person, i.e. addressed to "you", but there is also an option to specify another person. A value of `1` or `"I"` means you'll be addressed in the first person:

```{r}
how_far(due_date = due_date, person = 1)
```

Any other character string will be interpreted as a third-person name, which is useful, for example, if you're following the pregnancy progress of a partner:

```{r}
date_when(33, due_date = due_date, person = "Ruth")
```

## Tracking medications

For those who get pregnant via fertility treatment, it is likely they need to take a number of different medications to support the pregnancy. Having functionality for tracking these is useful for both practical and emotional reasons. When I was pregnant, in my personal, private pregnancy package, whenever I called my version of `how_far`, it would print a message with how many injections I had already endured, and how many I had left to go, which helped me get through them.

I haven't written that functionality into `how_far()` in this generalised version of the pregnancy package, but I have provided a separate `medications_remaining()` function to keep track of medications. It requires a data frame of medications, which must have the following columns with the specified data types:

| Column name | Data type | Description |
|----|----|----|
| medication | character or factor | name of the medication |
| format | character or factor | format of the medication (e.g. pill, injection) |
| quantity | numeric | number of units to take per day |
| start_date | Date | date to start taking the medication |
| stop_date | Date | final date on which medication is taken |

Note that if the quantity of a given medication changes during the pregnancy, 
you need separate rows with the start and stop dates for each quantity.

Here's an example of what a small `medications` table might look like 
(a larger example table is provided in the package as `pregnancy::medications`):

```{r}
# a simplified medication schedule
meds <- dplyr::tribble(
  ~medication, ~format, ~quantity, ~start_date, ~stop_date,
  "progynova", "tablet", 3, "2025-08-21", "2025-08-31",
  "progynova", "tablet", 6, "2025-09-01", "2025-09-11",
  "cyclogest", "pessary", 2, "2025-09-03", "2025-09-11",
  "clexane", "injection", 1, "2025-09-08", "2025-11-05"
) |>
  dplyr::mutate(
    start_date = as.Date(start_date),
    stop_date = as.Date(stop_date)
  )
```

You can then calculate the quantity of medications left to take grouped by either by medication (default) or format. By default, the calculation is for today (i.e. `on_date` = `Sys.Date()`). The resulting table assumes that the function is being called first thing in the day, i.e. before any of `on_date`'s medications have been taken. 

```{r}
medications_remaining(meds)
```

```{r}
medications_remaining(meds, group = "format")
```

You can specify a value other than today for `on_date`, as well as an `until_date`.
This is useful if you need (as I did) to order a couple of week's medication at the time,
and have to tell the pharmacy exactly what you need:

```{r}
medications_remaining(
  meds,
  on_date = as.Date("2025-09-01"),
  until_date = as.Date("2025-09-14")
)
```

## Global options

It would be very tedious to have to enter a due date every time you call `how_far()` or `date_when()` over the course of a pregnancy, especially since that date is constant throughout. The same goes for the medications table required for `medications_remaining()`.

To avoid this, the pregnancy package makes use of **global options**, which can be set with the `set_*` family of functions (`set_due_date()`, `set_person()`, `set_medications()`). 
These functions only set the options for the current R session. 

To ensure the options persist across all R sessions, we suggest setting them in your `.Rprofile`, and give an example at the end of this vignette.

Global options can be retrieved with the `get_*` family of functions (`get_due_date()`, `get_person()`, `get_medications()`).

Any global option can be unset by calling its `set_*` function with the argument `NULL`.

### `pregnancy.due_date`

You can set the `pregnancy.due_date` option using the `set_due_date()` function.

When that option is set, if the `due_date` argument to `how_far()` or `date_when()` is `NULL` (the default), that option is retrieved.

`set_due_date()` only sets the due date for the current R session. To make this option persist (recommended), it can be added to your `.Rprofile`. The console output of `set_due_date()` provides guidance on how to do this.

```{r}
set_due_date(due_date)
```

Then `how_far()` can be called without needing to specify any arguments, and `date_when()` only needs the target week:

```{r}
how_far()
```

To check what the option is set to, use `get_due_date()`.

### `pregnancy.person`

`set_person()` sets the global option `pregnancy.person` for the current R session. 
If this option is set it will be retrieved to specify the value of `person` in `how_far()` and `date_when()`, 
unless it is overriden by passing another value directly to the `person` argument. If the option is not set, the default second-person will be used.

```{r}
set_person(1)
how_far()
set_person(NULL)
how_far()
```

### `pregnancy.medications`

`set_medications()` sets the global option `pregnancy.medications` for the current R session. 

When that option is set, if the `meds` argument to `medications_remaining()` is `NULL` (the default), that option is retrieved.

```{r}
set_medications(pregnancy::medications)
medications_remaining()
```

### An example `.Rprofile`

Taking the advice to specify the options in your `.Rprofile`, here's an example of what that might look like:

``` r
options(
  pregnancy.due_date = as.Date("2025-09-16"),
  pregnancy.person = "I", # addressed in first person
  pregnancy.medications = dplyr::tribble(
    ~medication, ~format, ~quantity, ~start_date, ~stop_date,
    "progynova", "tablet", 3, as.Date("2025-04-21"), as.Date("2025-04-30"),
    "progynova", "tablet", 6, as.Date("2025-05-01"), as.Date("2025-07-11"),
    "cyclogest", "pessary", 2, as.Date("2025-05-03"), as.Date("2025-07-11"),
    "clexane", "injection", 1, as.Date("2025-05-08"), as.Date("2025-09-05")
  )
)
```

Note that it is best to avoid creating R objects in your ".Rprofile", e.g. a `medications` data frame outside of the call to `options()`, otherwise that object will be loaded into your global environment at the start of every R session.
